Compositional Certified Resource Bounds

### 2014-07-02 paper/main.tex
bugfix in the experiments
add speed results
more meaningless results
more work on section 5

\section{Automatic Analysis}

@@

The most interesting rules are {\sc Q:IncP}, {\sc Q:DecP}, and {\sc
  Q:Inc}.

The case of the update statement is more interesting and specific to
linear potential functions.  In the {\sc Q:IncP} rule the program
updates a variable $x$ to add it $y \ge 0$.  Since $x$ is changed, the
quantitative annotation must be updated to reflect the change of the
program state.  We write $x'$ the value of $x$ after the assignment
($x + y$). Since $x$ is the only variable changed, only intervals of
the form $\inter u x$ and $\inter x u$ will be resized.  Note that for
any $u$, $\inter x u$ will get smaller with the update, and if $x' \in
\inter x u$ we have $|\inter x u| = |\inter x {x'}| + |\inter {x'}
u|$.  But $|\inter x {x'}| = |\inter 0 y|$, this means that the
potential $q'_{0y}$ in the postcondition can be increased by $q_{xu}$
under the guard that $x' \in \inter x u$.  Dually, the interval
$\inter v x$ can get bigger with the update.  And we know that
$|\inter v {x'}| \le y + |\inter v x|$, so we decrease the potential
of $\inter 0 y$ by $q_{vx}$ to pay for this change.  Remark that the
rule forces it only for $v \not\in \mathcal U$, because if not, we
know that $x \le v$ so $|\inter v x| = 0$.

Another very technical rule is {\sc Q:Call}.  It needs to
account for the stack ``switching'' caused by the function
call, the argument/return value passing, and the preservation
of local variables.  We can sum up the requirements for
this rule as follows,
\begin{itemize}
\item
  any potential in the function pre/postcondition is equalized
  to its matching potential in the call pre/postcondition,
\item
  the potential of intervals $|\inter x y|$ is preserved
  if $x$ and $y$ are local variables,
\item
  and the unknown potentials after the call (e.g.
  $|\inter x g|$, with $x$ local and $g$ global)
  are zeroed out in the postcondition.
\end{itemize}
If $x$ and $y$ are local variables and $\code{f}(x,y)$ is
called, the rule takes care to split the potential of
$|\inter x y|$ in two parts, one part to perform the
computation in the function $f$ and one part to keep for
later use after the function call.  This splitting is
realized by the equations $Q = P + S$ and $Q' =
P' + S'$.  Arguments in the function precondition
$(\Gamma_f, Q_f)$ are named using a fixed vector $\Vargs$
of names different from all program variables.  This
prevents name conflicts from happening and ensures that
the substitution $[\Vargs/\vec x]$ is meaningful.
Symmetrically, we use the unique name $\Vret$ to represent
the return value in the function postcondition
$(\Gamma'_f, Q'_f)$.

All rules except of {\sc Q:Weak} and {\sc Q:Relax} are syntax
directed.

@@

The soundness of the analysis leverages the quantitative logic.
We propose to translate judgements of~\pref{fig:auto} into
quantitative triples.  We define a translation function
$\tr$, such that if a judgement $J$ in the automatic analysis
is derivable, $\tr(J)$ is derivable in the quantitative
logic.  Using this translated judgement and the soundness
of our logic it is possible to get a certified resource bound
for the analyzed program.

The translation of an assertion $(\Gamma, Q)$ in the automatic
analysis is defined by
$$
  \tr(\Gamma, Q) := \lambda\state .\, (\Gamma(\state)) + \Phi_Q(\state),
$$
where we write $\Phi_Q$ for the unique linear potential function
defined by the quantitative annotation $Q$.  We also need to
translate the assumptions in function contexts of the automatic
analysis, this is done as follows
%
\begin{align*}
\tr(\Gamma_f, Q_f, \Gamma'_f, Q'_f) :=
&\forall z \, \vec v \, v . (
  \lambda \_ \, \vec v \, \state .\, \tr(\Gamma_f, Q_f)(\state[\Vargs \mapsto \vec v]), \\
&\phantom{\forall z \, \vec v \, v . (}
  \lambda \_ \, v \, \state .\, \tr(\Gamma'_f, Q'_f)(\state[\Vret \mapsto v])
  ).
\end{align*}
%
These definitions let us translate the judgement $J = B; R; P \vdash
S \dashv P'$ in the context $\Delta$ by
$$
  \tr(J) :=
  (\lambda f .\, \tr(\Delta(f))); \tr(B); \tr(R) \vdash
    \htriple{\tr(P)}{S}{\tr(P')}.
$$
The soundness of the automatic analysis can now be stated
formally with the following theorem.
%
\begin{theorem}[Soundness of the analysis]
  If $J$ is a judgement derived with the rules of
  \pref{fig:auto}, then $\tr(J)$ is a quantitative
  Hoare triple derivable with the rules of
  \pref{fig:logic}.
\end{theorem}
%
\noindent
The proof of this theorem is constructive and basically
maps each rule of the automatic analysis directly to
its counterpart in the quantitative logic.  With the
computational content of the proof, we have an
effective algorithm to construct proof objects for
upper bounds derived automatically.

\begin{figure*}[t]
\begin{mathpar}
\Rule{Q:Skip}
{}
{ B; R; (\Gamma, Q) \vdash \code{skip} \dashv (\Gamma, Q) }
\and \Rule{Q:Break}
{}
{  (\Gamma, Q_B); R; (\Gamma, Q_B {+} M_b) \vdash \code{break} \dashv (\Gamma', Q') }
\and \Rule{Q:Return}
{ P = Q_R[\Vret/x]  
\\ \forall i \in \dom{P} . \, p_i = q_i
}
{ B; (\Gamma,Q_R); (\Gamma, Q_R{+} M_r) \vdash \code{return}~x \dashv (\Gamma',Q') }
\and \Rule{Q:Assert}
{}
{ B; R; (\Gamma, Q {+} M_a) \vdash \code{assert}~e \dashv (\Gamma \land e, Q) }
\and \Rule{Q:Tick}
{}
{ B; R; (\Gamma, Q{+}M_t(n)) \vdash \code{tick}(n) \dashv (\Gamma, Q) }
\\ \Rule{Q:Loop}
{ (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q {+} M_l) }
{ B; R; (\Gamma, Q) \vdash \code{loop}~S \dashv (\Gamma', Q') }
\and \Rule{Q:Update}
{ q'_{xy}, q'_{yx} \in \Qplusz
\\ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
}
{ B; R; (\Gamma[x/y], Q {+} M_u {+} M_e(y)) \vdash x \gets y \dashv (\Gamma, Q') }
\and \Rule{Q:IncP}
{ \Gamma \models 0 \le y
\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\}
\\ \textstyle q'_{0y} = q_{0y}
      + \sum_{u \in \mathcal U} q_{xu}
      - \sum_{v \not\in \mathcal U} q_{vx}
}
{ \textstyle
  B; R; (\Gamma[x/x {+} y], Q {+} M_u {+} M_e(x {+} y)) \vdash x \gets x + y \dashv (\Gamma, Q')
}
\and \Rule{Q:DecP}
{ \Gamma \models 0 \le y
\\ \mathcal U = \left\{ u \mid \Gamma \models x - y \in \inter u x \right\}
\\ \textstyle q'_{y0} = q_{y0}
      + \sum_{u \in \mathcal U} q_{ux}
      - \sum_{v \not\in \mathcal U} q_{xv}
}
{ \textstyle
  B; R; (\Gamma[x/x {-} y],Q {+} M_u {+} M_e(x {-} y)) \vdash x \gets x - y \dashv (\Gamma, Q')
}
\and \Rule{Q:Inc}
{ M = M_u + M_e(x {\pm} y)
\\\\ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
}
{ \textstyle
  B; R; (\Gamma[x/x {\pm} y], Q {+} M) \vdash x \gets x \pm y \dashv (\Gamma, Q')
}
\and \Rule{Q:If}
{ B; R; (\Gamma \land e, Q {-} M_c^1) \vdash S_1 \dashv (\Gamma',Q')
\\\\ B; R; (\Gamma \land \neg e, Q {-} M_c^2) \vdash S_2 \dashv (\Gamma',Q')
}
{ B; R; (\Gamma, Q{+}M_e(e)) \vdash \code{if}(e)~S_1~\code{else}~S_2 \dashv (\Gamma',Q') }
\and\Rule{Q:Seq}
{  B; R; (\Gamma, Q) \vdash S_1 \dashv (\Gamma', Q' {+} M_s)
\\\\ B; R; (\Gamma', Q') \vdash S_2 \dashv (\Gamma'',Q'')
}
{ B; R; (\Gamma, Q) \vdash S_1;S_2 \dashv (\Gamma'',Q'') }
\and \Rule{Q:Weak}
{ B; R; (\Gamma_2,Q_2) \vdash S \dashv (\Gamma'_2, Q'_2)
\\\\ \Gamma_1 \models \Gamma_2
\\ Q_1 \succeq_{\Gamma_1} Q_2
\\ \Gamma'_2 \models \Gamma'_1
\\ Q'_2 \succeq_{\Gamma_2'} Q'_1
}
{ B; R; (\Gamma_1,Q_1) \vdash S \dashv (\Gamma'_1, Q'_1) }
\and\Rule{Q:Call}
{ (\Gamma_f, Q_f,  \Gamma_f', Q_f') \in \Delta(f)
\and \text{Loc} = \text{Locals}(Q)
\and \forall i \neq j . \, x_i \neq x_j
\and c \in \Qplusz
\\ Q = P + S
\and Q' = P' + S
\and U = Q_f[\Vargs / \vec x]
\and U' = Q'_f[\Vret/r]
\\ \forall i \in \dom{U} . \, p_i = u_i
\and \forall i \in \dom{U'} .\, p'_i = u'_i 
\and \forall i \not\in \dom{U'} .\, p'_i = 0
\and \forall i \not\in \text{Loc}
  .\, s_i = 0
}
{ B; R; (\Gamma_f[\Vargs / \vec x] \land \Gamma_{\text{Loc}}, Q {+} c {+} M_f)
  \vdash r \gets f(\vec x) \dashv
  (\Gamma_f'[\Vret/r] \land \Gamma_{\text{Loc}}, Q' {+} c {-} M_r)
}
\and \Rule{Relax}
{ \mathcal L = \{ xy \mid \exists l_{xy} {\in} \N \, . \, \Gamma \models l_{xy} \le |\inter x y| \}
\and \mathcal U = \{ xy \mid \exists u_{xy}{\in} \N \, . \, \Gamma \models |\inter x y| \le u_{xy} \}
\\ \forall i \in \mathcal U.\, q'_i \ge q_i - r_i
\\ \forall i \in \mathcal L.\, q'_i \ge q_i + p_i
\\ \forall i \not\in \mathcal U {\cup} \mathcal L {\cup} \{0\} .\, q'_i \ge q_i
\\ q'_0 \geq q_0 {\textstyle + \sum_{i \in \mathcal U} u_i r_i - \sum_{i \in \mathcal L} l_i p_i}
}
{ Q'   \succeq_\Gamma Q }
\and \Rule{Q:Extend}
{ \fenv f = (\vec y, S_f)
\\ B; (\Gamma_f', Q'_f); (\Gamma_f[\Vargs/\vec y], Q_f[\Vargs/ \vec y] ) \vdash S_f \dashv (\Gamma', Q')
}
{ (\Gamma_f, Q_f, \Gamma_f', Q_f') \in \Delta(f) }
\end{mathpar}
\caption{Inference rules of the quantitative analysis}
\label{fig:auto}
\end{figure*}

\section{Soundness Theorems}

The soundness of a judgement relates the cost semantics with the potential
functions derived using our proof system.  Depending on the outcome of
the semantics, we give different soundness statements.
But, before tackling the issue of soundness let us remind and define some
convenient notations.
\begin{itemize}
\item $H \mapsto \Phi_H(Q)$ for $Q$ a quantitative context, is the
unique linear potential function that has $Q(i)$ as a coefficient for the
base function $f_i$.
\item $H \models \Gamma$ is true if and only if all the assertions in
$\Gamma$ are true on the heap $H$.
\end{itemize}

Ideally we would
like our judgements to be meaningful even in the case of a program
crash.  More precisely, we want the first potential function to be enough
to pay for the execution of the program up to the crash point.  This is
formally stated in the following theorem.

\begin{theorem}[Busy case]
For any $S$, $H$, $q$, $q'$, $\Gamma$, $\Gamma'$, $Q$ and $Q'$
such that $H \models \Gamma$, $\Gamma; Q \vdash S \dashv Q'; \Gamma'$,
  and $(H, S) \Downarrow_{(q,q')} \circ$,
we have $q \le \Phi_H(Q)$.
\end{theorem}
\begin{proof}
Simple application of the terminating case proved afterwards.
\end{proof}

In order to prove the harder case for terminating programs, need to first
focus on the only structural rule of the proof system: {\sc Q:Weak}.  This
rule can be applied anywhere in the derivation, it is not syntax directed.
To prove the soundness of this rule, we need a lemma that gives a
semantic explanation of the $\succeq_\Gamma$ relation.

\begin{lemma}[Relax]
If $Q' \succeq_\Gamma Q$, for any heap $H$ such that $H \models \Gamma$,
we have $\Phi_H(Q') \ge \Phi_H(Q)$.
\end{lemma}
\begin{proof}
We observe the following inequations.
\begin{align*}
\Phi_H(Q') &= \sum_i q'_i f_i(H) \\
&\ge q_0 + \sum_{i\in\mathcal U} u_i p_i - \sum_{i\in\mathcal L} l_i p_i
  + \sum_{i\in\mathcal U} (q_i - p_i) f_i(H) + \sum_{i\in\mathcal L} (q_i + p_i) f_i(H)
  + \sum_{i\not\in \mathcal U \cup \mathcal L} q_i \\
 &\ge \Phi_H(Q)
  + \sum_{i\in\mathcal U} (u_i - f_i(H)) p_i
  + \sum_{i\in\mathcal L} (f_i(H) - l_i) p_i \\
 &\ge \Phi_H(Q).
\end{align*}
Since $H \models \Gamma$, using the transitivity of $\models$ and
the definition of $\mathcal U$ and $\mathcal L$ we get
$\forall i\in\mathcal L.\, l_i \le f_i(H)$ and
$\forall i\in\mathcal U.\, f_i(H)\le u_i$.
These two inequalities justify the last step.
\end{proof}

Using this lemma, we can tackle the hardest theorem of soundness.  This theorem
needs a strong statement to make the proof by induction go through and defines
the complete meaning of a judgement $\Gamma; Q \vdash S \dashv Q'; \Gamma'$.
In a similar fashion than the busy case, the initial quantitative context is enough
to perform the whole computation: $q \le \Phi_H(Q)$.  But additionally, the
two quantitative contexts are also related to the resource consumption $q - q'$
of the piece of program executed.  This second requirement exactly matches
the semantic condition described in the overview section: it says that the
initial potential $\Phi_H(Q)$ is enough to both pay for the resource
consumption of the program $q - q'$ and pay for the potential of the resulting
heap $\Phi_{H'}(Q')$.

The proof of this result if very technical but basically goes through by induction
on the big-step evaluation.  Only a representative subset of all the cases
is present in the following proof.

\begin{theorem}[Terminating case]
For any $S$, $H$, $H'$, $q$, $q'$, $\Gamma$, $\Gamma'$, $Q$ and $Q'$ such
that $\Gamma; Q \vdash S \dashv Q'; \Gamma'$ and $H \models \Gamma$ and
$(H,s) \Downarrow_{(q,q')} H'$, we have
$H' \models \Gamma'$ and
$q \le \Phi_H(Q)$ and
$\Phi_H(Q) - \Phi_{H'}(Q') \ge q - q'$.
\end{theorem}

\begin{proof} By nested induction on the big step semantics and on the
judgement derivation (to account for the structural rule {\sc Q:Weak}).

>>>>

\section{Automatic Amortized Analysis}

@@

Most interesting are the rules {\sc Q:IncP}, {\sc Q:DecP}, and {\sc
  Q:Inc} for increments and decrements, which describe how the
potential is distributed after a size change of the variables.  The
rule {\sc Q:IncP} is for increments $x \gets x + y$ in which we can
deduce from the logical context $\Gamma$ that $y$ is not negative.
Similarly, {\sc Q:DecP} is for increments $x \gets x - y$ in which we
can deduce that $y \ge 0$.  Of course, we have symmetrical rules {\sc
  Q:IncN} and {\sc Q:DecN} that can be applied if $y$ is not positive.
The rules are equivalent in the case where $y=0$.  Finally, the {\sc
  Q:Inc} can be applied if we cannot deduce any information about $y$.
In the implementation, the rules for increments and decrements are
combined into one syntax directed rule where the specialized rules
take priority over {\sc Q:Inc}.

To explain how rules for increment and decrement work, it is
sufficient to understand the rule {\sc Q:IncP}.  The other follow the
same idea and are symmetrical.  In {\sc Q:IncP}, the program updates a
variable $x$ with $x+y$ where $y \ge 0$.  Since $x$ is changed, the
quantitative annotation must be updated to reflect the change of the
program state.  We write $x'$ the value of $x$ after the assignment.
Since $x$ is the only variable changed, only intervals of the form
$\inter u x$ and $\inter x u$ will be resized.  Note that for any $u$,
$\inter x u$ will get smaller with the update, and if $x' \in \inter x
u$ we have $|\inter x u| = |\inter x {x'}| + |\inter {x'} u|$.  But
$|\inter x {x'}| = |\inter 0 y|$ which means that the potential
$q'_{0y}$ in the postcondition can be increased by $q_{xu}$ under the
guard that $x' \in \inter x u$.  Dually, the interval $\inter v x$ can
get bigger with the update.  We know that $|\inter v {x'}| \le y +
|\inter v x|$.  So we decrease the potential of $\inter 0 y$ by
$q_{vx}$ to pay for this change.  Remark that the rule insures this
only for $v \not\in \mathcal U$ because we know that $x \le v$
otherwise, and thus $|\inter v x| = 0$.

Another interesting rule is {\sc Q:Call}.  It needs to account for the
changes to the stack caused by the function call, the argument/return
value passing, and the preservation of local variables.  We can sum up
the main ideas of the rule as follows.
\begin{itemize}
\item The potential in the pre- and postconditions of the function
  specification is equalized to its matching potential in the callee's
  pre/postcondition.
\item
  The potential of intervals $|\inter x y|$ is preserved
  if $x$ and $y$ are local variables,
\item
  The unknown potentials after the call (e.g.
  $|\inter x g|$, with $x$ local and $g$ global)
  are set to zero in the postcondition.
\end{itemize}
If $x$ and $y$ are local variables and $\code{f}(x,y)$ is
called, the rule takes care to split the potential of
$|\inter x y|$ in two parts, one part to perform the
computation in the function $f$ and one part to keep for
later use after the function call.  This splitting is
realized by the equations $Q = P + S$ and $Q' =
P' + S'$.  Arguments in the function precondition
$(\Gamma_f, Q_f)$ are named using a fixed vector $\Vargs$
of names different from all program variables.  This
prevents name conflicts from happening and ensures that
the substitution $[\Vargs/\vec x]$ is meaningful.
Symmetrically, we use the unique name $\Vret$ to represent
the return value in the function postcondition
$(\Gamma'_f, Q'_f)$.

The rule {\sc Q:Weak} is not syntax directed.  In the implementation
we apply {\sc Q:Weak} before loops and between the two statements of a
sequential composition.  We could integrate weakening into every
syntax directed rule but this simple heuristic helps to make the
analysis efficient.  The high-level idea of {\sc Q:Weak} is the same
as in the rule {\sc L:Weak} of the quantitative logic: If we have a
sound judgement, then it is sound to add more potential to the
precondition and remove potential from the postcondition.  The concept
of \emph{more potential} is formalized by the relation $Q'
\succeq_\Gamma Q$ that is defined in the rule {\sc Relax}.  Here, $Q$
and $Q'$ are potential annotations and $\Gamma$ is a logical context.
The relax rule deals also with the important task of transferring
constant potential (represented by $q_0$) to interval sizes and vise
versa.  If we can deduce from the logical context that the interval
size $|\inter x y| \geq l$ is larger then a constant $l$ then we can
transfer potential $q_{xy} {\cdot} |\inter x y|$ form the interval to
constant potential $l{\cdot}q_{xy}$ and guarantee that we do not gain
potential.  Conversely, if $|\inter x y| \leq u$ for a constant $u$
then we can transfer constant potential $u{\cdot}q_{xy}$ to the
interval potential $q_{xy} {\cdot} |\inter x y|$ without gaining
potential.

\paragraph{Automatic Inference via LP Solving.}

We separate the search of a derivation into two steps.  As a first
step we go through the whole program and apply inductively the
inference rules of the automatic amortized analysis.  During this
process our tool uses symbolic names for the coefficients $(q_i)$.
Every time a linear constraint must be satisfied by these
coefficients, it is recorded using the symbolic names in a global
list.  When the list of all constraints that must be satisfied by the
coefficients is collected, we feed it to an off the shelf
LP-solver~\footnote{We currently use Coin-Or CLP.}.  If the solver
successfully finds a solution to the given problem, we know that a
derivation exists for the considered program.  We can then extract the
initial $Q$ from the solver and get a resource bound for the program.
To get a full derivation we simply extract the complete solution from
the solver, and apply it to the symbolic names $(q_i)$ of the
coefficients in the derivation.  If the LP-solver fails to find a
solution, an error is reported to the user.

@@

The soundness of the analysis builds on the quantitative logic.  We
translate judgements of the automatic amortized analysis as defined in
\pref{fig:auto} into quantitative Hoare triples.  We define a
translation function $\tr$, such that if a judgement $J$ in the
automatic analysis is derivable, $\tr(J)$ is derivable in the
quantitative logic.  By using $\tr$ to translate derivations of the
automatic analysis to derivations in the quantitative logic we can
automatically obtain a certified resource bound for the analyzed
program.

The translation of an assertion $(\Gamma, Q)$ in the automatic
analysis is defined by
$$
  \tr(\Gamma, Q) := \lambda\state .\, (\Gamma(\state)) + \Phi_Q(\state),
$$
where we write $\Phi_Q$ for the unique linear potential function
defined by the quantitative annotation $Q$.  We also need to
translate the assumptions in function contexts of the automatic
analysis.  This is done as follows.
%
\begin{align*}
\tr(\Gamma_f, Q_f, \Gamma'_f, Q'_f) :=
&\forall z \, \vec v \, v . (
  \lambda \_ \, \vec v \, \state .\, \tr(\Gamma_f, Q_f)(\state[\Vargs \mapsto \vec v]), \\
&\phantom{\forall z \, \vec v \, v . (}
  \lambda \_ \, v \, \state .\, \tr(\Gamma'_f, Q'_f)(\state[\Vret \mapsto v])
  )
\end{align*}
%
These definitions let us translate the judgement $J = B; R; P \vdash
S \dashv P'$ in the context $\Delta$ by
$$
  \tr(J) :=
  (\lambda f .\, \tr(\Delta(f))); \tr(B); \tr(R) \vdash
    \htriple{\tr(P)}{S}{\tr(P')}.
$$
The soundness of the automatic analysis can now be stated
formally with the following theorem.
%
\begin{theorem}[Soundness of the analysis]
  If $J$ is a judgement derived with the rules of
  \pref{fig:auto}, then $\tr(J)$ is a quantitative
  Hoare triple derivable with the rules of
  \pref{fig:logic}.
\end{theorem}
%
\noindent
The proof of this theorem is constructive and basically maps each rule
of the automatic analysis directly to its counterpart in the
quantitative logic.  The most tricky parts are the translations of the
rules for increments and decrements and the rule {\sc Q:Weak} for
weakening because we have to show that the preconditions of the rules
{\sc L:Weak} or {\sc L:Update}, respectively, are met.  For instance
we prove the following lemma to show the soundness of the translation
of {\sc Q:Weak}.

\begin{lemma}[Relax]
  If $Q' \succeq_\Gamma Q$ and $\state$ is a program state such that
  $\state \models \Gamma$ then we have $\Phi_{Q'}(\state) \ge \Phi_Q(\state)$.
\end{lemma}
\begin{proof}
We observe the following inequations.
\begin{align*}
\Phi_{Q'}(\state) &= \sum_i q'_i f_i(H) \\
&\ge q_0 + \sum_{i\in\mathcal U} u_i p_i - \sum_{i\in\mathcal L} l_i p_i
  + \sum_{i\in\mathcal U} (q_i - p_i) f_i(bH)  \\ & \;\;\;\;\;\;\;\; + \sum_{i\in\mathcal L} (q_i + p_i) f_i(H)
  + \sum_{i\not\in \mathcal U \cup \mathcal L} q_i \\
 &\ge \Phi_Q(\state)
  + \sum_{i\in\mathcal U} (u_i - f_i(H)) p_i
  + \sum_{i\in\mathcal L} (f_i(H) - l_i) p_i \\
 &\ge \Phi_Q(\state).
\end{align*}
Since $H \models \Gamma$, using the transitivity of $\models$ and
the definition of $\mathcal U$ and $\mathcal L$ we get
$\forall i\in\mathcal L.\, l_i \le f_i(H)$ and
$\forall i\in\mathcal U.\, f_i(H)\le u_i$.
These two inequalities justify the last step.
\end{proof}
%
\noindent
We prove similar lemmas for the rules that deal with assignment.  With
the computational content of the proof, we have an effective algorithm
to construct certificates for upper bounds derived automatically.

\begin{figure*}[t]
\begin{mathpar}
%
\Rule{Q:Skip}
{}
{ B; R; (\Gamma, Q) \vdash \code{skip} \dashv (\Gamma, Q) }
%
\and \Rule{Q:Break}
{}
{  (\Gamma, Q_B); R; (\Gamma, Q_B {+} M_b) \vdash \code{break} \dashv (\Gamma', Q') }
%
\and \Rule{Q:Return}
{ P = Q_R[\Vret/x]  
\\ \forall i \in \dom{P} . \, p_i = q_i
}
{ B; (\Gamma,Q_R); (\Gamma, Q_R{+} M_r) \vdash \code{return}~x \dashv (\Gamma',Q') }
%
\and \Rule{Q:Assert}
{}
{ B; R; (\Gamma, Q {+} M_a) \vdash \code{assert}~e \dashv (\Gamma \land e, Q) }
%
\and \Rule{Q:Tick}
{}
{ B; R; (\Gamma, Q{+}M_t(n)) \vdash \code{tick}(n) \dashv (\Gamma, Q) }
%
\\ \Rule{Q:Loop}
{ (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q {+} M_l) }
{ B; R; (\Gamma, Q) \vdash \code{loop}~S \dashv (\Gamma', Q') }
%
\and \Rule{Q:Update}
{ q'_{xy}, q'_{yx} \in \Qplusz
\\ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
}
{ B; R; (\Gamma[x/y], Q {+} M_u {+} M_e(y)) \vdash x \gets y \dashv (\Gamma, Q') }
%
\and \Rule{Q:IncP}
{ \Gamma \models y \ge 0
\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\}
\\ \textstyle q'_{0y} = q_{0y}
      + \sum_{u \in \mathcal U} q_{xu}
      - \sum_{v \not\in \mathcal U} q_{vx}
}
{ \textstyle
  B; R; (\Gamma[x/x {+} y], Q {+} M_u {+} M_e(x {+} y)) \vdash x \gets x + y \dashv (\Gamma, Q')
}
%
\and \Rule{Q:DecP}
{ \Gamma \models y \ge 0
\\ \mathcal U = \left\{ u \mid \Gamma \models x - y \in \inter u x \right\}
\\ \textstyle q'_{y0} = q_{y0}
      + \sum_{u \in \mathcal U} q_{ux}
      - \sum_{v \not\in \mathcal U} q_{xv}
}
{ \textstyle
  B; R; (\Gamma[x/x {-} y],Q {+} M_u {+} M_e(x {-} y)) \vdash x \gets x - y \dashv (\Gamma, Q')
}
%
\and \Rule{Q:Inc}
{ M = M_u + M_e(x {\pm} y)
\\\\ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
}
{ \textstyle
  B; R; (\Gamma[x/x {\pm} y], Q {+} M) \vdash x \gets x \pm y \dashv (\Gamma, Q')
}
%
\and \Rule{Q:If}
{ B; R; (\Gamma \land e, Q {-} M_c^1) \vdash S_1 \dashv (\Gamma',Q')
\\\\ B; R; (\Gamma \land \neg e, Q {-} M_c^2) \vdash S_2 \dashv (\Gamma',Q')
}
{ B; R; (\Gamma, Q{+}M_e(e)) \vdash \code{if}(e)~S_1~\code{else}~S_2 \dashv (\Gamma',Q') }
%
\and\Rule{Q:Seq}
{  B; R; (\Gamma, Q) \vdash S_1 \dashv (\Gamma', Q' {+} M_s)
\\\\ B; R; (\Gamma', Q') \vdash S_2 \dashv (\Gamma'',Q'')
}
{ B; R; (\Gamma, Q) \vdash S_1;S_2 \dashv (\Gamma'',Q'') }
%
\and \Rule{Q:Weak}
{ B; R; (\Gamma_2,Q_2) \vdash S \dashv (\Gamma'_2, Q'_2)
\\\\ \Gamma_1 \models \Gamma_2
\\ Q_1 \succeq_{\Gamma_1} Q_2
\\ \Gamma'_2 \models \Gamma'_1
\\ Q'_2 \succeq_{\Gamma_2'} Q'_1
}
{ B; R; (\Gamma_1,Q_1) \vdash S \dashv (\Gamma'_1, Q'_1) }
%
\and\Rule{Q:Call}
{ (\Gamma_f, Q_f,  \Gamma_f', Q_f') \in \Delta(f)
\and \text{Loc} = \text{Locals}(Q)
\and \forall i \neq j . \, x_i \neq x_j
\and c \in \Qplusz
\\ Q = P + S
\and Q' = P' + S
\and U = Q_f[\Vargs / \vec x]
\and U' = Q'_f[\Vret/r]
\\ \forall i \in \dom{U} . \, p_i = u_i
\and \forall i \in \dom{U'} .\, p'_i = u'_i 
\and \forall i \not\in \dom{U'} .\, p'_i = 0
\and \forall i \not\in \text{Loc}
  .\, s_i = 0
}
{ B; R; (\Gamma_f[\Vargs / \vec x] \land \Gamma_{\text{Loc}}, Q {+} c {+} M_f)
  \vdash r \gets f(\vec x) \dashv
  (\Gamma_f'[\Vret/r] \land \Gamma_{\text{Loc}}, Q' {+} c {-} M_r)
}
%
\and \Rule{Q:Extend}
{ \fenv f = (\vec y, S_f)
\\ B; (\Gamma_f', Q'_f); (\Gamma_f[\Vargs/\vec y], Q_f[\Vargs/ \vec y] ) \vdash S_f \dashv (\Gamma', Q')
}
{ (\Gamma_f, Q_f, \Gamma_f', Q_f') \in \Delta(f) }
%
\and \Rule{Relax}
{ \mathcal L = \{ xy \mid \exists l_{xy} {\in} \N \, . \, \Gamma \models l_{xy} \le |\inter x y| \}
\and \mathcal U = \{ xy \mid \exists u_{xy}{\in} \N \, . \, \Gamma \models |\inter x y| \le u_{xy} \}
\\ \forall i \in \mathcal U.\, q'_i \ge q_i - r_i
\\ \forall i \in \mathcal L.\, q'_i \ge q_i + p_i
\\ \forall i \not\in \mathcal U {\cup} \mathcal L {\cup} \{0\} .\, q'_i \ge q_i
\\ q'_0 \geq q_0 {\textstyle + \sum_{i \in \mathcal U} u_i r_i - \sum_{i \in \mathcal L} l_i p_i}
}
{ Q'   \succeq_\Gamma Q }

\end{mathpar}
\caption{Inference rules of the quantitative analysis}
\label{fig:auto}
\end{figure*}

% \section{Implementation}

% \subsection{Logical Annotations Generation}

% \paragraph{Program conditions.}
% The first pass in our OCaml implementation generates the
% logical annotations $\Gamma$ around all program statements.
% These annotations are used later on by the quantitative
% analysis to generate contraints.
% %
% The first design choice we had to make is to give a fixed form
% to program conditions.  Because of their generality and because
% they are closed under negation, we decided to consider
% inequalities over linear combinations as program conditions.
% \begin{mathpar}
% C := L < L \mid L > L \mid L \le L \mid L \ge L
% \and
% L := x \mid k \mid k * L \mid L + L \mid L - L
% \end{mathpar}
% A Condition is represented internally as a pair of one constant
% $k \in \mathbb N$ and one map $m : V \rightarrow \mathbb N$.
% Its intended semantics is
% $$
% \llbracket m \rrbracket(H) = k + \sum_v m_v \cdot H(v) \le 0.
% $$
% Note that conditions of this shape are closed by negation, as showed
% by the following equivalence
% $$
% \neg (k + \sum_v m_v \cdot H(v) \le 0)
% \Leftrightarrow (1-k) + \sum_v -m_v \cdot H(v) \le 0
% $$
% We can also note that all the ``questions'' asked to the logic
% assertions during a proof derivation can be described using
% linear combinations.  Conveniently enough, this language of
% assertions is exactly Presburger's arithmetic.  This means that
% we have a sound and complete decision procedure to answer
% all the questions asked during the analysis.

% \paragraph{Logical contexts.}
% The second problem related to the logic is how to figure out
% the logical contexts $\Gamma$ for each program statement.
% This is a question that is answered by abstract interpretation
% but, to show that our work really does not need the full power
% of abstract interpretation, we designed and implemented a very
% simple top-down approach: starting from a statement $S$ and
% a pre-condition $\Gamma$ (given as a list of inequalities), we
% generate $\mathcal P(S, \Gamma)$ a postcondition for $S$ using the
% following equations:
